---
title: "Principle Component Analysis for World Environmental Data"
author: "Conner Smith"
date: "2/12/2022"
output: 
  html_document: 
    theme: flatly
    code_folding: hide
---

```{r setup, include=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(ggfortify)
library(cowplot)
library(kableExtra)
library(janitor)
```

## **Overview**

This analysis looks at several key country level environmental variables to determine relationships between these variables. Specifically, crop cover, tree cover, rainfall, elevation, and temperature and weather patterns are considered to establish this assessment. This was compiled from the Google Earth Engine which consolidates publicly available remote sensing data.   

## **Analysis**

```{r}
# Read in the data 

world_env <- read_csv(here('data', 'world_env_vars.csv'))
```

The following presents the results of a principle components analysis on several selected environmental variables by country. It includes bith a biplot to show the relationship between variables and a screeplot showing the amount of variance captured in each principle component. 

```{r}
# Filter the data to set up the PCA 

world_clean <- world_env %>% 
  column_to_rownames('Country') %>% 
  select(elevation, cropland_cover, 
         tree_canopy_cover, rain_mean_annual,
         temp_mean_annual, cloudiness) %>%
  clean_names('title') %>% # Generates nice figure labels 
  drop_na()

# Now conduct the PCA 

world_pca <- world_clean %>% 
  scale() %>% 
  prcomp()

pca_summary <- summary(world_pca)
# pca_summary

# 90% of variance covered in first 4 components 
```

#### **Figure 1: World Environmental Data Variable Exploration**
```{r, fig.height= 7, fig.width= 7}
# Build the biplot using the 'autoplot function.

biplot<- autoplot(world_pca,
         data = world_clean,
         colour = 'blue',
         alpha = 0.5,
         loadings = TRUE,
         loadings.label = TRUE,
         loadings.colour = 'black',
         loadings.label.colour = 'black',
         loadings.label.size = 3,
         loadings.label.vjust = -0.7) +
  theme_minimal()

# Create the screeplot manually for more customization 

sd_vec <- world_pca$sdev

var_vec <- (sd_vec)^2

pc_names <- colnames(world_pca$rotation)

pct_expl_df <- data.frame(v = var_vec,
                          pct_v = var_vec/ sum(var_vec),
                          pc = fct_inorder(pc_names)) %>% 
  mutate(pct_label = paste0(round(pct_v * 100, 1), '%'))

# Create an object for the plot 

scree <- ggplot(pct_expl_df, aes(x = pc, y = v)) +
  geom_col() +
  geom_text(aes(label = pct_label,
                vjust = 0, nudge_y = 0.005)) +
              labs(x = 'Principle Component',
                   y = 'Variance Estimated')

#Combine the plots using 'cowplot'

plot_grid(biplot, scree,
                   ncol = 1,
                   rel_heights = c(1.5, 1))


```

##### **Figure 1:** *This figure presents a biplot and screeplot afrom a principle components analysis on the a range of environemntal variables by country.*

From **Figure 1**, we can determine the following:

- More than 90 percent of variance is accounted for in the first four principle components.

- Tree canopy cover and average annual rainfall are closely correlated. 

- Cloudiness and average annual temperature seem to have no relationship.

- Cropland cover and annual mean temperature seem to have a negative relationship. 

- Cropland cover and elevation are clsoely correlated. 

### **Data Citation**

*Venter, Z. and W.G. Whitford. 2018. Environmental variables for world countries. Kaggle. https://www.kaggle.com/zanderventer/environmental-variables-for-world-countries/metadata*